---
title: "Assignment 2: Coding Basics"
author: "Isabel Zungailia"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. # Generated a sequence of numbers from 1 to 100, increasing by fours. Assigned sequence the name "s".
seq(1, 100, by = 4)
s <- seq(1, 100, by = 4)

#2. Calculated the mean of "s". Gave the mean value the name of "mean.result".
mean (s)
mean.result <- mean(s)
    # Calculated the median of "s". Gave the median value the name of "median.result".
median (s)
median.result <- median (s)

#3. Asked R to determine if the mean is greateer than the median. 
is_mean_greater <- mean.result > median.result
print(is_mean_greater)
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5. Created three vectors, each with four components. Titled the vectors "names", "scores", and "pass_or_fail".
names <- c("Isabel", "Libby", "Jenna", "Jaimie")

scores <- c(98, 95, 95, 90)

pass_or_fail <- c(TRUE, TRUE, TRUE, TRUE)

#6. Labeled each vector with a comment on what type of vector it is. 
names # character vector
scores # numerical vector
pass_or_fail # logical vector

#7. Combined the three vectors into a data frame. Named the data frame "student_scores".
student_scores <- data.frame(names=names, scores=scores, pass_or_fail=pass_or_fail)
student_scores

#8. Labeled the data frame columns with the following names: "Student Name", "Test Score", and "Pass?". 
colnames(student_scores) <- c("Student Name", "Test Score", "Pass?")
student_scores
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: # A data frame is a two-dimensional object that has rows and columns, essentially a data table. A matrix is similar to a dataframe, except all values in a matrix are numeric. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
#10. Used an 'ifelse' statement to determine whether a given test score is a passing grade of 50 or above.
ifelse(student_scores$`Test Score` > 50, print("Passed"), print("Failed"))

grade_function <- function (scores) {
  ifelse(student_scores$`Test Score` > 50, print("Passed"), print("Failed"))
}

#11. Applied 'grade_function' to 'scores' vector from #5. 
grade_function(scores)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: # When I tried to create an 'if' and 'else' statement, I got an error because the condition length is greater than 1. 'If' and 'else' is used when testing a single condition. The 'ifelse' statement worked because it is able to evaluate all the elements in the vector without producing an error. 
