---
title: "Assignment 09: Data Scraping"
author: "Isabel Zungailia"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1.
#Check working directory
getwd()

#Load necessary packages
library(tidyverse)
library(lubridate)
library(rvest)
library(dplyr)

#Set theme
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2.Use rvest's `read_html()` function to bring the contents of the website into our coding environment
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3.
#Set the element address variables
water.system.name_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
pswid_tag <- 'td tr:nth-child(1) td:nth-child(5)'
ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
max.withdrawals.mgd_tag <- 'th~ td+ td'

#Scrape the data items
water.system.name <- webpage %>% html_nodes(water.system.name_tag) %>% html_text()
pswid <- webpage %>%   html_nodes(pswid_tag) %>%  html_text()
ownership <- webpage %>% html_nodes(ownership_tag) %>% html_text()
max.withdrawals.mgd <- webpage %>% html_nodes(max.withdrawals.mgd_tag) %>% html_text()

#Check values
water.system.name
pswid
ownership
max.withdrawals.mgd
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4. Convert scraped data into a dataframe
df_withdrawals <- data.frame(Month = as.factor(c("Jan", "May", "Sept", "Feb", "June", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"))) %>% 
  mutate(Water_System_Name = !!water.system.name,
         PSWID = !!pswid,
         Ownership = !!ownership,
         Maximum_Withdrawals_MGD = !!max.withdrawals.mgd,
         Month = factor(Month, levels = month.name),
         Date = my(paste(Month,"-",Year))) %>% 
  arrange(Month)

df_withdrawals2 <- data.frame(Month_abbr = c("Jan", "May", "Sept", "Feb", "June", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"), Month = c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12), Year = rep(2021,12), Max_withdrawals_mgd = as.numeric(max.withdrawals.mgd))

df_withdrawals2 <- df_withdrawals2 %>% 
  mutate(Water_System_Name = !!water.system.name,
         PSWID = !!pswid,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year))) %>% 
  arrange(Month)

#5. Create a line plot of the maximum daily withdrawals across the months for 2021
ggplot(df_withdrawals2,aes(x=Month_abbr,y=Max_withdrawals_mgd)) +
  geom_line(aes(group=1)) + 
  geom_smooth(method="loess",se=FALSE) +
 labs(title = paste("2021 Water Usage Data For",water.system.name),
       y="Maximum Daily Withdrawals (mgd)",
       x="Month") %>% 

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021')

the_url <- ifelse(the_year==2021,
                    'https://www.ncwater.org/WUDC/app/LWSP/report.php?',
                    paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?',pwsid,'&',Year))

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
webpage2 <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2015')



```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8


```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9


```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?