---
title: "Assignment 5: Data Visualization"
author: "Isabel Zungailia"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 14th @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy

[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the

[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd() #Check working directory
library(tidyverse) #Load the tidyverse package
library(lubridate) #Load the lubridate package
library(cowplot) #Load the cowplot package

#Upload processed data files
PeterPaul.chem.nutrients <- 
read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv",
           stringsAsFactors = TRUE)

NiwotRidge.litter <-
  read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv",
           stringsAsFactors = TRUE)

#2 
#Change format of dates to 'Date' for both datasets
class(PeterPaul.chem.nutrients$sampledate)
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
class(PeterPaul.chem.nutrients$sampledate)

class(NiwotRidge.litter$collectDate)
NiwotRidge.litter$collectDate <- as.Date(NiwotRidge.litter$collectDate, format = "%Y-%m-%d")
class(NiwotRidge.litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
#Build a theme
mytheme <- theme_classic(base_size = 12) + #Set text size to 12
  theme(axis.text = element_text(color = "black"), #Change text color to black
        legend.position = "top") #Set the legend to be positioned at the top of the plot

#Set the theme as the default for subsequent plots
theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4
#Plot total phosphorus (`tp_ug`) by phosphate (`po4`)
totalphos.by.phosphate <-
  ggplot(PeterPaul.chem.nutrients, aes(x = po4, y = tp_ug)) + 
  xlab(expression("Phosphate, po4 (µg/l)")) + #Change x-axis label
  ylab(expression("Total Phosphorus, tp_ug (µg/l)")) + #Change y-axis label
  xlim(0, 40) + #Adjust axis to hide extreme values
  ylim(0, 150) +
  guides(color=guide_legend(title="Lake Name")) + #Change legend title to "Lake Name"
  guides(shape=guide_legend(title="Lake Name")) +
  geom_point(aes(color = lakename, shape = lakename)) + #Separate aesthetics for Peter and Paul lakes
  geom_smooth(method = lm, color = "black") #Add a line of best fit, color it black
print(totalphos.by.phosphate)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5
#Make three separate boxplots of (a) temperature, (b) TP, and (c) TN
#Temperature boxplot
boxplotA <- #Convert month integers into factors and include factors 1-12 in the output
  ggplot(PeterPaul.chem.nutrients, aes(x = factor(month, levels = c(1:12)), y = temperature_C)) + 
  xlab(expression("Month")) + #Set axis labels
  ylab(expression("Temperature (C)")) +
  guides(color=guide_legend(title="Lake Name")) + #Set legend title to "Lake Name
  geom_boxplot(aes(color = lakename)) + #Set lake as a color aesthetic
    scale_x_discrete(drop = FALSE) #Prevent ggplot from dropping levels having no data
print(boxplotA)

#Total phosphorous boxplot
boxplotB <- #Convert month integers into factors and include factors 1-12 in the output
  ggplot(PeterPaul.chem.nutrients, aes(x = factor(month, levels = c(1:12)), y = tp_ug)) + 
  xlab(expression("Month")) + #Set axis labels
  ylab(expression("Total Phosphorus, tp_ug (µg/l)")) +
  guides(color=guide_legend(title="Lake Name")) + #Set legend title to "Lake Name
  geom_boxplot(aes(color = lakename)) + #Set lake as a color aesthetic
  scale_x_discrete(drop = FALSE) #Prevent ggplot from dropping levels having no data
print(boxplotB)

#Total nitrogen boxplot
boxplotC <- #Convert month integers into factors and include factors 1-12 in the output
  ggplot(PeterPaul.chem.nutrients, aes(x = factor(month, levels = c(1:12)), y = tn_ug)) + 
  xlab(expression("Month")) + #Set axis labels
  ylab(expression("Total Nitrogen, tn_ug (µg/l)")) +
  guides(color=guide_legend(title="Lake Name")) + #Set legend title to "Lake Name"
  geom_boxplot(aes(color = lakename)) + #Set lake as a color aesthetic
  scale_x_discrete(drop = FALSE) #Prevent ggplot from dropping levels having no data
print(boxplotC)

#Make a cowplot with the three graphs - boxplotA, boxplotB, boxplotC
#Make sure there is only one legend and that the axis are aligned
combined_plotsABC <-
plot_grid(boxplotA + theme(legend.position = "none"), 
          boxplotB +  theme(legend.position = "none"), 
          boxplotC + theme(legend.position = "bottom"), 
          nrow = 1, align = 'hv', axis = "bt", rel_heights = c(1.6, 1.7, 2.3))

print(combined_plotsABC)

```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: The variables of interest - Temperature (C), Total Phosphorous (µg/l), and Nitrogen (µg/l) - follow relatively similar trends in both lakes throughout the year. The mean temperature was slightly higher in Paul Lake throughout every month except October and November, and both lakes recorded their highest temperatures in the months of August and September. The mean total phosphorous (µg/l) levels were slightly higher in Peter Lake throughout every month that was sampled. The greatest mean total phosphorous for Paul Lake was recorded in the months of May and September, while the greatest mean total phosphorous for Peter Lake was recorded in September. The mean nitrogen (µg/l) levels were also slightly higher in Peter Lake (than Paul) throughout every month that was sampled. The greatest mean nitrogen level for Paul Lake was recorded in the month of September, while the greatest mean nitrogen count for Peter Lake was recorded in August. Overall, it appears that the months with warmer temperatures generate higher levels of total phosphorous and nitrogen.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6
#Plot dry mass of needle litter by date
drymass_by_date <-
  ggplot(NiwotRidge.litter[NiwotRidge.litter$functionalGroup %in% "Needles",], 
         aes(x = collectDate, y = dryMass)) + #Display the "Needles" functional group
  ylab(expression("Dry Mass (g)")) +
  guides(color=guide_legend(title="NLCD Class")) + #Set legend title
  guides(shape=guide_legend(title="NLCD Class")) +
  geom_point(aes(color = nlcdClass, shape = nlcdClass)) #Separate by NLCD class with a color aesthetic
  print(drymass_by_date)

#7
#Plot dry mass of needle litter by date and separate NLCD classes into three facets
drymass_by_date2 <-
  ggplot(NiwotRidge.litter[NiwotRidge.litter$functionalGroup %in% "Needles",], 
         aes(x = collectDate, y = dryMass)) + #Display the "Needles" functional group
  geom_point() +
  facet_wrap(vars(nlcdClass), nrow = 3) + #Separate NLCD classes into three facets
    xlab(expression("Date")) + #Set axis labels
  ylab(expression("Dry Mass (g)"))
  print(drymass_by_date2)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: After comparing the plots (6 vs. 7), I think the plot in #7 is more effective in showing the comparison of trends between the three NLCD classes in the "Needles" functional group. The 'facet_wrap' function generate multi-panel plots, which was a better way to display this dataset rather than having all the classes mixed together (as in #6). It is helpful to be able to visualize the individual trends of dry mass in each NLCD class, but then we can also make comparsions between all three classes by layering the individual plots on top of each other (with 'facet_wrap'). 