---
title: "Assignment 2: Coding Basics"
author: "Isabel Zungailia"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
seq(1, 100, by = 4)
s <- seq(1, 100, by = 4)

#2. 
mean (s)
mean.result <- mean(s)

median (s)
median.result <- median (s)

#3. 
is_mean_greater <- mean.result > median.result
print(is_mean_greater)
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5. 
names <- c("Isabel", "Libby", "Jenna", "Jaimie")

scores <- c(98, 95, 95, 90)

pass_or_fail <- c(TRUE, TRUE, TRUE, TRUE)

#6.
names # character vector
scores # numerical vector
pass_or_fail # logical vector

#7.
student_scores <- data.frame(names=names, scores=scores, pass_or_fail=pass_or_fail)

#8. 
colnames(student_scores) <- c("Student Name", "Test Score", "Pass?")

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: # A data frame is a two-dimensional object that has rows and columns, essentially a data table. A matrix is similar to a dataframe, except all values in a matrix are numeric. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
#9. 

#10. 
if("scores" > 50) {
  print("Passing grade")
} else {
  print("Failed")
}

ifelse(student_scores$`Test Score` > 50, print("Passed"), print("Failed"))


#11. 
score_i <- 98
score_l <- 95
score_jenna <- 95
score_jaimie <- 90

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: #
