---
title: "Assignment 5: Data Visualization"
author: "Isabel Zungailia"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 14th @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd() #Check working directory
library(tidyverse) #Load the tidyverse package
library(lubridate) #Load the lubridate package
library(cowplot) #Load the cowplot package

#Upload processed data files
PeterPaul.chem.nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

NiwotRidge.litter <-
  read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 
#Change format of dates to 'Date' for both datasets
class(PeterPaul.chem.nutrients$sampledate)
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
class(PeterPaul.chem.nutrients$sampledate)

class(NiwotRidge.litter$collectDate)
NiwotRidge.litter$collectDate <- as.Date(NiwotRidge.litter$collectDate, format = "%Y-%m-%d")
class(NiwotRidge.litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3 Build a theme - set text size to 14, changed text color to black, and set the legend position to be at the top of the plot
mytheme <- theme_classic(base_size = 12) + 
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

#Set the theme as the default for subsequent plots
theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4 Plot total phosphorus (`tp_ug`) by phosphate (`po4`)
totalphos.vs.phosphate <-
  ggplot(PeterPaul.chem.nutrients, aes(x = po4, y = tp_ug)) + 
    xlim(0, 50) + #Adjust axis to hide extreme values
  ylim(0, 150) +
    geom_point(aes(color = lakename)) + #Separate aesthetics for Peter and Paul lakes
  geom_smooth(method = lm) #Add a line of best fit
print(totalphos.vs.phosphate)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5
# Make three separate boxplots of (a) temperature, (b) TP, and (c) TN
boxplotA <-
  ggplot(PeterPaul.chem.nutrients, aes(x = factor(month, levels = c(1:12)), y = temperature_C)) +
  geom_boxplot(aes(color = lakename)) +
  theme(legend.position = "none") +
    scale_x_discrete(drop = FALSE)
print(boxplotA)

boxplotB <-
  ggplot(PeterPaul.chem.nutrients, aes(x = factor(month, levels = c(1:12)), y = tp_ug)) +
  geom_boxplot(aes(color = lakename)) +
  theme(legend.position = "none") +
  scale_x_discrete(drop = FALSE)
print(boxplotB)

boxplotC <-
  ggplot(PeterPaul.chem.nutrients, aes(x = factor(month, levels = c(1:12)), y = tn_ug)) +
  geom_boxplot(aes(color = lakename)) +
  theme(legend.position = "none") +
  scale_x_discrete(drop = FALSE)
print(boxplotC)

#install.packages("cowplot")
library(cowplot)

combined_plots <-
  plot_grid(boxplotA, boxplotB, boxplotC, align = 'h', axis = "bt")

legend <- get_legend(boxplotA +
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom"))
  
  plot_grid(combined_plots, legend, nrow = 2, align = 'h', axis = "bt")
  

```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6
drymass_by_date <-
  ggplot(NiwotRidge.litter[NiwotRidge.litter$functionalGroup %in% "Needles",], aes(x = collectDate, y = dryMass)) +
  geom_point(aes(color = nlcdClass))
  print(drymass_by_date)

#7
drymass_by_date2 <-
  ggplot(NiwotRidge.litter[NiwotRidge.litter$functionalGroup %in% "Needles",], aes(x = collectDate, y = dryMass)) +
  geom_point()
  print(drymass_by_date2)


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: