---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Isabel Zungailia"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1.
#Check working directory
getwd()
#Load necessary packages
library(tidyverse)
library(lubridate)
#If needed, install/update package "htmltools" before installing and loading agricolae
#install.packages("htmltools")
library(htmltools)
#install.packages("agricolae")
library(agricolae)

#Import the raw NTL-LTER raw data file for chemistry/physics
NTL_LTER.RAW.chem.phys <-
  read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",stringsAsFactors = TRUE)

#Set date columns to date objects
class(NTL_LTER.RAW.chem.phys$sampledate)
NTL_LTER.RAW.chem.phys$sampledate <- as.Date(NTL_LTER.RAW.chem.phys$sampledate, format = "%m/%d/%y")
class(NTL_LTER.RAW.chem.phys$sampledate)

#2.
#Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no correlation between lake temperature (C) and depth (slope and y-intercept are 0).
Ha: There is a correlation between lake temperature (C) and depth. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4.
#Use a pipe function to wrangle the NTL-LTER dataset
NTL_LTER.RAW.chem.phys.wrangled <-
  NTL_LTER.RAW.chem.phys %>%
    mutate(month = month(sampledate)) %>% 
      filter(month == 7) %>%  #Select only dates in July
        select(lakename:daynum, depth:temperature_C) %>% #Select specific columns
          na.omit() #Omit NAs so that only complete cases exists

#5.
temperature.by.depth <-
  ggplot(NTL_LTER.RAW.chem.phys.wrangled, aes(x = depth, y = temperature_C)) + 
  xlab(expression("Depth (m)")) + #Change x-axis label
  ylab(expression("Temperature (C)")) + #Change y-axis label
  ylim(0, 35) +  #Adjust axis to hide extreme values
  geom_point() +
  geom_smooth(method = lm, color = "black") #Add a line of best fit
print(temperature.by.depth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This figure suggests that temperature (C) decreases as depth increases (m). The Distribution of points suggest...


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
temperature_depth.regression <- lm(data = NTL_LTER.RAW.chem.phys.wrangled, depth ~ temperature_C)
summary(temperature_depth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
The residuals refer to the 
The r-squared value is used to understand how close the data set fits into the model, and describes the percent of variance accounted for by the explanatory vairables (correlation squared; ranges in value from 0-1). In our model, the r-squared value is 0.7387, which means that the model closely decsribes the variance between the two (73.87% of temeprature values are explained by depth).
The p-value is 2.2e-16, which is very close to zero and indicates that there is a strong relationship between depth and temperature (we can reject the null hypothesis).


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
NTL_LTER_AIC <-
  lm(data = NTL_LTER.RAW.chem.phys.wrangled, year4 ~ temperature_C)
summary(NTL_LTER_AIC)
step(NTL_LTER_AIC)
 #remove year4?

#10
predicting_temp_regression <-
  lm(data = NTL_LTER.RAW.chem.phys.wrangled, daynum ~ temperature_C)
summary(predicting_temp_regression)
step(predicting_temp_regression)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
July_Lake_Temps.anova <- #One-way ANOVA test using 
  aov(data = NTL_LTER.RAW.chem.phys.wrangled, temperature_C ~ lakename)
summary(July_Lake_Temps.anova)

July_Lake_Temps.anova2 <-
  lm(data = NTL_LTER.RAW.chem.phys.wrangled, temperature_C ~ lakename)
summary(July_Lake_Temps.anova)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The results of the ANOVA tests indicated that there is a signifcant difference in the mean temperature (C) among the lakes. The p-value from the ANOVA is <2e-16 (much less than 0.05), which means that there are significant differences between the groups in question.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

```

>Answer: 
